\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{titlesec}
\usepackage{setspace}

\setlength{\parskip}{0.3em}
\setlength{\parindent}{0pt}

\titlespacing*{\section}
{0pt}{0.6em}{0.3em}

\begin{document}

\begin{center}
    {\Large \textbf{ECE51216: Digital Systems Design Automation}}\\
    {\normalsize Spring 2026}\\[0.4em]
    {\Large \textbf{Intermediate Project Report}}\\[0.3em]
    {\normalsize Option 1: Build}\\[0.6em]
    \textbf{Soomin Lee, Dogyu Ryu}
\end{center}

\section{Project Overview}

Our project follows Option 1 from the proposal: building a SAT solver based on the DPLL
(Davis--Putnam--Logemann--Loveland) framework for CNF formulas in DIMACS format. The
original goal was to start from a correct baseline DPLL solver and then extend it with more
efficient data structures and heuristics, especially watched literals and VSIDS. The intermediate
goal for this phase was therefore to establish a working baseline implementation that can parse
benchmark instances, perform recursive search, and produce measurable statistics for later
comparison.

\section{Progress Since the Proposal}

The main progress so far has been the implementation of a fresh Python baseline solver rather than
continuing directly from older reference code. This decision was made to keep the code structure
simple, readable, and aligned with our own understanding of the algorithm. The current solver has
already reached the stage where it can read benchmark files, perform search, and report basic
execution statistics. In addition, we organized a small self-contained benchmark set inside the
repository so that the project can be run and demonstrated without relying on external folder
dependencies.

The implemented components are listed below.

\begin{itemize}
    \item \textbf{DIMACS parser.} We implemented input parsing for CNF problems in DIMACS
    format. The parser handles comment lines, problem headers, standard clause termination with
    \texttt{0}, and SATLIB-style end markers.
    \item \textbf{Baseline DPLL recursion.} The solver performs recursive branching on
    unassigned literals and backtracks when a contradiction is found.
    \item \textbf{Boolean Constraint Propagation (BCP).} Before each branching step, the solver
    repeatedly detects unit clauses and propagates forced assignments.
    \item \textbf{Formula simplification.} After each assignment, clauses satisfied by that literal
    are removed, while falsified literals are deleted from the remaining clauses.
    \item \textbf{Instrumentation.} The implementation records the number of decisions,
    propagations, conflicts, backtracks, recursive calls, maximum recursion depth, and runtime.
\end{itemize}

At this stage, the emphasis has been on correctness and modularity rather than optimization. This
is intentional, because the baseline version will serve as the control point when we later compare
the effect of watched literals and VSIDS.

\section{Current Implementation Status}

The current solver is able to classify both satisfiable and unsatisfiable instances correctly on the
sample benchmark inputs we tested. In addition, it can optionally print a satisfying assignment in a
DIMACS-style literal format when a SAT instance is solved. At the current stage, the repository
contains a local benchmark set of seven CNF instances: three SAT cases and four UNSAT cases.
This benchmark set is not intended to be the final evaluation suite, but it is sufficient to verify the
baseline solver and produce an initial intermediate report with reproducible results.

To validate the baseline implementation, we ran the solver on all seven local benchmark inputs:

\begin{center}
\small
\renewcommand{\arraystretch}{1.1}
\begin{tabular}{|p{3.0cm}|p{1.2cm}|p{1.2cm}|p{1.5cm}|p{2.1cm}|p{2.4cm}|}
\hline
\textbf{Instance} & \textbf{Type} & \textbf{Result} & \textbf{Decisions} & \textbf{Propagations} & \textbf{Runtime(s)} \\
\hline
\texttt{example-1.cnf} & SAT & SAT & 4 & 13 & 0.000253 \\
\hline
\texttt{uf20-01.cnf} & SAT & SAT & 24 & 42 & 0.000340 \\
\hline
\texttt{uf20-04.cnf} & SAT & SAT & 22 & 67 & 0.000553 \\
\hline
\texttt{unsat.cnf} & UNSAT & UNSAT & 2 & 2 & 0.000007 \\
\hline
\texttt{unsat1.cnf} & UNSAT & UNSAT & 6 & 4 & 0.000017 \\
\hline
\texttt{unsat2.cnf} & UNSAT & UNSAT & 0 & 3 & 0.000007 \\
\hline
\texttt{unsat3.cnf} & UNSAT & UNSAT & 30 & 101 & 0.000232 \\
\hline
\end{tabular}
\end{center}

The results above are preliminary and were collected primarily to validate the end-to-end baseline
pipeline from parsing to recursive search. However, they already provide some useful observations.
First, the solver handled both SAT and UNSAT instances correctly on the local set. Second, the
instances vary enough to show different search behaviors. For example, \texttt{unsat2.cnf} is
resolved entirely by propagation without any branching decisions, while \texttt{unsat3.cnf}
requires substantially more decisions and propagations than the smaller UNSAT cases. Among the
SAT instances in the current set, \texttt{uf20-01.cnf} and \texttt{uf20-04.cnf} already produce
nontrivial backtracking behavior, making them reasonable early targets for comparing future
heuristics.

We also verified that the solver can emit a concrete satisfying assignment when a SAT instance is
found. For example, on \texttt{example-1.cnf}, the solver produced a valid DIMACS-style model in
addition to the SAT/UNSAT classification and execution statistics. This is useful both for debugging
and for demonstrating correctness in the report.

\section{Changes From the Original Proposal}

The core project objective has not changed: we are still building a SAT solver using the DPLL
framework and planning to evaluate the effect of algorithmic heuristics on performance. However,
two practical changes have been made since the proposal.

First, we decided to use \textbf{Python instead of C/C++} for the implementation. The main reason
is faster prototyping and easier debugging during the development phase. Since the course project
focuses more on algorithmic understanding and comparison than on absolute industrial-level
performance, Python is sufficient for implementing and evaluating the required ideas.

Second, the implementation order has been slightly adjusted. Instead of trying to introduce several
advanced heuristics at once, we first focused on producing a correct and instrumented baseline.
This staged approach makes the later evaluation cleaner, because we will be able to compare:

\begin{itemize}
    \item baseline DPLL,
    \item DPLL with watched literals,
    \item DPLL with watched literals and VSIDS.
\end{itemize}

This comparison structure is consistent with the original evaluation plan, but the development has
been reorganized to prioritize a reliable baseline first.

Finally, we made a small reproducibility-oriented change in project organization. Instead of
depending on an external reference directory for every test input, we copied a compact benchmark
set directly into the project repository. This change does not affect the technical goals of the
proposal, but it makes the current implementation easier to run, share, and evaluate.

\section{Remaining Work and Plan}

The next implementation step is to replace naive clause scanning with a watched-literal data
structure so that propagation becomes more efficient. After that, we will add the VSIDS decision
heuristic to improve branching quality. Once these features are integrated, we will run a more
systematic benchmark evaluation using CNF instances from SATLIB and other DIMACS-style test
sets.

Our remaining work is summarized as follows.

\begin{itemize}
    \item Implement watched literals and verify that propagation remains correct.
    \item Add the VSIDS heuristic for decision variable selection.
    \item Expand the benchmark set to include both SAT and UNSAT cases of varying sizes.
    \item Compare solver versions using runtime, number of decisions, number of propagations,
    conflicts, and backtracks.
    \item Summarize the results and discuss which heuristic contributes the most to performance.
\end{itemize}

\section{Conclusion}

In summary, the project has progressed from planning to a working Python baseline SAT solver.
The current implementation already supports DIMACS parsing, DPLL recursion, unit propagation,
and execution statistics, and it has been validated on several sample SAT and UNSAT instances.
Although watched literals and VSIDS are not yet implemented, the project is now at a stable point
where these extensions can be added and evaluated in a structured way.

\end{document}
